---
title: "BDBV 2026 outbreak: phylogenetic dating analysis"
author: "Vinicius Franceschi and Erik Volz"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
# Force knitr to use the R session's working directory)
knitr::opts_knit$set(root.dir = here::here())
```

```{css, echo=FALSE}
/* Default scroll for all outputs */
pre {
  max-height: 300px;
  overflow-y: auto;
}

/* Taller box for verbose outputs */
.scroll-300 {
  max-height: 300px;
  overflow-y: auto;
}
```

# Summary

This report analyses genomic data generated by other researchers and originally described in the [Virological.org post](https://virological.org/t/genomic-epidemiology-of-the-ongoing-2026-bundibugyo-virus-disease-outbreak-in-the-democratic-republic-of-the-congo/1045). There is no intention to publish these analyses. Instead, this report provides complementary analyses to help understand the ongoing Bundibugyo ebolavirus (BDBV) outbreak.

## Data download and curation

The sequence alignment and associated metadata were downloaded from Pathoplexus on **9 July 2026**. The original post describes **139** sequences sampled between **2 May 2026** and **23 June 2026**; however, **149** sequences were available on Pathoplexus at the time of download.

Before analysis, we performed several curation steps, largely following the recommendations in the original post:

1. Removed sequence **PP_00764QW** because of evidence of ADAR editing.
2. Trimmed the alignment to **18,900 nt** to remove sequencing artefacts at the 3' end (introduced by sequence **PP_006XHL9.3**, which was retained).
3. Masked the first **65 nt** and last **432 nt**, as these positions were covered in only a small number of sequences.

To make the analyses as comparable as possible with the original report, we downloaded the corresponding [Pathoplexus SeqSet](https://pathoplexus.org/seqsets/PP_SS_2740.1) and matched sequence identifiers. This yielded **127** matching sequences, indicating that the datasets are unfortunately not identical. In addition, sequence **PP_00711T3** failed the IQ-TREE composition test and was removed prior to downstream analyses.

## Alignment and tree reconstruction

Following curation, the Pathoplexus alignment was aligned to the **NC_014373** 2007 BDBV reference genome using MAFFT with the `--keeplength` and `--add` options. A maximum-likelihood (ML) phylogeny was then reconstructed using IQ-TREE 2 with ModelFinder, which selected the **K3Pu+F** substitution model (`-m MFP -B 1000 --bnni -alrt 1000`).

Quality-control checks using [Raccoon](https://github.com/artic-network/raccoon) (`aln-qc` and `tree-qc`) did not identify any issues requiring further action.

All subsequent analyses were based on this ML tree and included:

- data loading and tree rooting;
- molecular clock dating using multiple models in **treedater** and **BactDating**;
- identification and removal of temporal outliers using **treedater** and **DiagnoDating**;
- molecular clock model comparison and selection;
- root-to-tip regression;
- estimation of uncertainty in evolutionary rates and node dates;
- estimation of effective population size through time; and
- reconstruction of geographic transitions using maximum parsimony.

## 1. Load libraries

```{r libraries}
library(here)
library(ape)
library(phytools)
library(glue)
library(dplyr)
library(treedater) # devtools::install_github(...)
library(lubridate)
library(ggtree) # from Bioconductor
library(ggplot2)
library(DiagnoDating, quietly = TRUE)
```

## 2. Load tree & metadata

### 2.1 Tree & alignment settings

```{r load-tree}
set.seed(9876)

TR_DIR <- "results/2026_07_09_5pm"
tr <- read.tree(glue("{TR_DIR}/bdbv.treefile"))
tr

# Alignment length after masking:
aln_len <- 18900 - 65 - 432
```

### 2.2 Metadata

```{r load-metadata}
DATA_DIR <- "data/2026_07_09_5pm/"
F_NAME <- "ebola-bdbv_metadata_2026-07-09T1550.tsv"

md <- read.csv(glue("{DATA_DIR}/{F_NAME}"), sep = "\t")

#colnames(md)
table(md$geoLocAdmin2)
```

### 2.3 Match metadata to tree tips

```{r match-metadata}
length(intersect(tr$tip.label, md$accessionVersion))

md_match <- md %>%
  filter(accessionVersion %in% tr$tip.label) %>%
  dplyr::slice(match(tr$tip.label, accessionVersion))

md_match$sampleCollectionDate <- as.Date(md_match$sampleCollectionDate)
md_match$num_date <- decimal_date(md_match$sampleCollectionDate)

# Named vector of sample times
sts <- md_match$num_date
names(sts) <- md_match$accessionVersion
```

---

## 3. Root tree & drop outgroup

```{r root-tree}
.rootanddrop <- function(tr0) {
  tr_rooted <- root(tr0, outgroup = "NC_014373", resolve.root = TRUE)
  drop.tip(tr_rooted, tip = "NC_014373")
}

tr <- .rootanddrop(tr)
```

### 3.1 Root & subsample bootstrap trees

```{r bootstrap-trees}
nboot <- 200
btres <- read.tree(glue("{TR_DIR}/bdbv.ufboot"))
btres <- btres[sample.int(length(btres), size = nboot, replace = FALSE)]
btres <- lapply(btres, .rootanddrop)
```

---

## 4. Initial clock dating (strict)

```{r initial-dating, class.output="scroll-300"}
td0 <- dater(
  tr, sts, s = aln_len,
  clock = "strict",
  meanRateLimits = c(0.0005, 0.002),
  numStartConditions = 20, quiet = FALSE,
  ncpu = 4
)

```

---

## 5. Outlier Removal

```{r outlier-removal}
.otips <- function(td) {
  ot <- outlierTips(td)
  ot[ot$q < 0.05, "taxon"]
}

ot <- .otips(td0)
td1 <- td0
tr1 <- tr

while (length(ot) > 0) {
  tr1 <- drop.tip(tr1, ot)
  td1 <- dater(
    tr1, sts, s = aln_len,
    clock = "strict",
    meanRateLimits = c(0.0005, 0.002),
    numStartConditions = 20, quiet = TRUE, #FALSE
    ncpu = 4
  )
  ot <- .otips(td1)
}

# Tips removed as outliers (empty = none removed)
print(setdiff(td0$tip.label, td1$tip.label))
```

---

## 6. Clock model selection

### 6.1 Relaxed clock test

```{r relaxed-clock-test, class.output="scroll-300"}
rct <- relaxedClockTest(
  tr1, sts, aln_len,
  meanRateLimits = c(0.0005, 0.002),
  quiet = FALSE, ncpu = 8, nreps=100,
  overrideTempConstraint = FALSE
)
```

### 6.2 Fit uncorrelated & additive relaxed clocks

```{r relaxed-clock-fits1, class.output="scroll-300"}
# Uncorrelated
td1u <- dater(
  tr1, sts, s = aln_len,
  clock = "uncorrelated",
  meanRateLimits = c(0.0005, 0.002),
  numStartConditions = 20, quiet = FALSE,
  ncpu = 4
)
```

```{r relaxed-clock-fits2, class.output="scroll-300"}
# Additive
td1a <- dater(
  tr1, sts, s  = aln_len,
  clock = "additive",
  meanRateLimits = c(0.0005, 0.002),
  numStartConditions = 20, quiet = FALSE,
  ncpu = 4
)

```

```{r relaxed-clock-fits3}
# Compare log-likelihoods: uncorrelated vs additive
c(td1u$loglik, td1a$loglik)
```

### 6.3 Inspect best-fit model (strict)

```{r inspect-additive1}
# Time tree plot
plot(td1, show.tip.label = F)
axisPhylo(root.time = td1$timeOf, backward = FALSE)
```

```{r inspect-additive2}
# Summary
td1

# MRCA date
date_decimal(td1$timeOfMRCA)
```

### 6.4 Root-to-tip regression

```{r root-to-tip}
rootToTipRegressionPlot(td1)
```

---

## 7. Bootstrap for confidence intervals (from iqtree ufboot distribution)

```{r bootstrap-dating, class.output="scroll-300"}
# Bootstrap
btd <- treedater::boot(
  td1, btres,
  ncpu = 8,
  overrideTempConstraint = FALSE, quiet = TRUE
)

btd
```
```{r bootstrap-dating2}
# Bootstrap CI on MRCA date
date_decimal(btd$timeOf)
```

---

## 8. BactDating

```{r bactdating}
trbd1 <- tr1
trbd1$edge.length <- trbd1$edge.length * aln_len

bd1a <- bactdate(
  trbd1, sts[tr1$tip.label],
  initMu = aln_len * td1$mean.rate,
  model = "arc", showProgress = FALSE # TRUE
)

bd1a
```

> **Note:** The estimated mean rate from BactDating is consistent with treedater, but the inferred root date may not be biologically plausible.


---

## 9. DiagnoDating

### 9.1 Root-to-tip (DiagnoDating)

NEEDS DiagnoDating >= 0.9.1, ie the fix-treedater-units branch (PR #1), which is where the
additive clock, the seqlen argument and matching of named dates were fixed. Before that,
treedater results always came back as model='poisson' with relax=0:
```
devtools::install_github('xavierdidelot/DiagnoDating', ref = 'fix-treedater-units')
```

```{r diagnodating-roottotip}
roottotip(tr1, sts)
```

DiagnoDating, like BactDating, wants branch lengths in substitutions rather than per site,
so rescale the tree. seqlen then tells treedater how to convert back, and omega0 and
meanRateLimits stay in treedater's own units of substitutions per site per year.

```{r diagnodating-convert}
trdd <- tr1
trdd$edge.length <- trdd$edge.length * aln_len
```

### 9.2 Run & diagnose

```{r diagnodating-run}
ddtd <- runDating( trdd, sts[ tr1$tip.label ], algo = 'treedater', keepRoot = TRUE, 
                   seqlen = aln_len, clock = 'strict', omega0 = td1$mean.rate, meanRateLimits = td1$mean.rate*c(1,1+1e-6) ) #additive
```

```{r diagnodating-plots}
ddtd
plotLikBranches(ddtd)
plotResid(ddtd)
```

```{r diagnodating-test}
# Anderson-Darling normality test on residuals
testResid(ddtd)
```

---

## 10. Profile likelihood for the clock rate

```{r lik-profile1, results='hide'}
ratetologlik <- function(omega){
	print( omega )
	td <- dater(tr1, sts, s=aln_len,  clock="additive", omega0 = omega, meanRateLimits = omega*c(.9,1.4), quiet=TRUE, ncpu=4)
	c( td$adjusted.mean.rate, td$loglik )
}

omegas <- seq( 0.0001, 0.002, length.out = 30)
llprof <- sapply( omegas, ratetologlik )
omegas <- llprof[1,]
llprof <- llprof[2,]
source( 'scripts/profplot-02b.R')
```
```{r lik-profile2}
prpl <- profplot( omegas, llprof )
prpl
```

---

## 11. Effective population size (mlesky)

```{r mlesky1}
library(mlesky)
range_tr <- max(sts) - td1$timeOfMRCA
range_tr_weeks <- ceiling(range_tr * 52.1775)
range_tr_weeks
```

```{r mlesky2, results='hide'}
mlsk <- mlskygrid(td1, sampleTimes=sts, res=NULL, tau=NULL, tau_lower=0.001, tau_upper=100, 
                  ncross=10, ncpu=4, quiet=F, model=2)
plot(mlsk, logy=T)
mlsk$res
mlsk$tau 
```

```{r mlesky3, results='hide'}
mlsk_pboot <- mlesky::parboot(mlsk, nrep=1000, ncpu=8, dd=F)
# regular bootstrap NEEDS mlesky >= 0.1.9: # devtools::install_github('emvolz-phylodynamics/mlesky')
# bmlsk <- mlesky::boot( mlsk, btd$trees, ncpu = 8, sampleTimes = sts )
plot(mlsk_pboot, logy=T, ggplot=T)
```

---

## 12. Ancestral state estimation with tree figure

```{r treefig}
library(phangorn)

collapse_dated <- function(td, aln_len, min_subs = 0.5) {
	phy  <- as.phylo(td)
	subs <- td$intree$edge.length * aln_len              # substitutions per edge (aligned to phy$edge)
	nd   <- node.depth.edgelength(phy)                   # node heights (time) from root
	tmp  <- phy; tmp$edge.length <- subs
	topo <- di2multi(tmp, tol = min_subs)                # collapse by substitutions
	Ntt  <- Ntip(topo); dts <- numeric(Ntt + topo$Nnode) # re-time from the original treedater dates
	dts[1:Ntt] <- (td$timeOfMRCA + nd[1:Ntip(phy)])[match(topo$tip.label, phy$tip.label)]
	desc <- Descendants(topo, (Ntt+1):(Ntt+topo$Nnode), "tips")
	for (i in seq_along(desc)) {
		tips <- topo$tip.label[desc[[i]]]
		mn   <- if (length(tips)==Ntip(phy)) Ntip(phy)+1 else getMRCA(phy, tips)
		dts[Ntt+i] <- td$timeOfMRCA + nd[mn]
	}
	topo$edge.length <- dts[topo$edge[,2]] - dts[topo$edge[,1]]
	topo$root.time   <- td$timeOfMRCA
	topo
}
.norm <- function(x){ x <- toupper(trimws(x)); x[x=="" | x=="NOT PROVIDED"] <- "UNKNOWN"
	dplyr::recode(x, "MONGWALU"="MONGBWALU","MUNGWALU"="MONGBWALU","NYAKUNDE"="NYANKUNDE") }
# shared locality set + palette (both trees have identical tips, so colours are comparable)
loc0 <- .norm(md_match$geoLocAdmin2[ match(as.phylo(td1)$tip.label, md_match$accessionVersion) ])
top5 <- names(sort(table(loc0), decreasing=TRUE))[1:5]
LEV  <- c(top5, "Other")
pal  <- setNames(c("#0072B2","#D55E00","#009E73","#E69F00","#CC79A7","grey72"), LEV)
mrsd <- as.Date(date_decimal(max(sts))); ci <- btd$timeOfMRCA_CI; break_gr <- 2026.35
mb <- decimal_date(seq(as.Date("2026-01-01"), as.Date("2026-08-01"), by="month"))
ml <- format(seq(as.Date("2026-01-01"), as.Date("2026-08-01"), by="month"), "%b %Y")

geo_tree_fig <- function(tree, root_date, title) {
	phy <- as.phylo(tree); nt <- Ntip(phy)
	loc <- .norm(md_match$geoLocAdmin2[ match(phy$tip.label, md_match$accessionVersion) ])
	loc[!loc %in% top5] <- "Other"
	dat <- phangorn::phyDat(as.matrix(data.frame(s=loc, row.names=phy$tip.label)), type="USER", levels=LEV)
	mat <- do.call(rbind, lapply(phangorn::ancestral.pars(phy, dat, "MPR"), function(z) z[1,])); colnames(mat) <- LEV
	map  <- LEV[apply(mat,1,which.max)]; conf <- apply(mat,1,max)
	ntrans <- sum(map[phy$edge[,1]] != map[phy$edge[,2]])
	lab <- rep(NA_character_, nt+phy$Nnode)               # UFBoot labels, only on collapsed trees
	if (!is.binary(phy)) {
		ufb <- as.numeric(sub(".*/", "", tr1$node.label))
		for (n in (nt+1):(nt+phy$Nnode)) {
			tips <- phy$tip.label[ unlist(phangorn::Descendants(phy, n, "tips")) ]
			mn <- if (length(tips)==nt) Ntip(tr1)+1 else getMRCA(tr1, tips)
			if (!is.null(mn) && !is.na(mn)) lab[n] <- as.character(round(ufb[mn - Ntip(tr1)]))
		}
	}
	nodedf <- data.frame(node=1:(nt+phy$Nnode), state=factor(map,levels=LEV), conf=conf, lab=lab)
	pg <- ggtree(tree, mrsd=mrsd, size=0.5) %<+% nodedf + aes(color=state)
	# ggtree rounds mrsd to whole days, shifting its x-axis; off re-aligns absolute-date annotations
	rootx <- pg$data$x[pg$data$node == nt+1]; rooty <- pg$data$y[pg$data$node == nt+1]
	off <- rootx - root_date; cih <- 0.4   # half-height of the thin TMRCA CI band, in tip-row units
	pg$layers <- c(annotate("rect", xmin=ci[1]+off, xmax=ci[2]+off, ymin=rooty-cih, ymax=rooty+cih, fill="grey30", alpha=0.45), pg$layers)  # thin TMRCA 95% CI band at the root, behind the tree
	pg +
		geom_vline(xintercept=break_gr+off, linetype="dashed", color="grey55", linewidth=0.4) +
		# annotate("label", x=break_gr+off, y=nt*0.965, label="growth-rate\nchange 2026.35", size=2.7,
		#          color="grey40", hjust=-0.04, vjust=1, lineheight=0.9, fill="white", alpha=0.7) +
		annotate("point", x=rootx, y=rooty, shape=18, size=2.8, color="grey20") +
		annotate("text", x=mean(ci)+off, y=rooty+cih+nt*0.02, label="TMRCA 95% CI", size=2.6, color="grey35", vjust=0) +
		geom_nodepoint(aes(color=state, size=conf), alpha=0.85) +
		geom_nodelab(aes(label=lab), geom="label", size=1.9, vjust=-0.6, color="grey20",
		             fill="white", label.size=0, label.padding=unit(0.04,"lines"), na.rm=TRUE) +
		geom_tippoint(aes(fill=state), shape=21, size=2, stroke=0.25, color="white") +
		scale_color_manual(values=pal, guide="none") +
		scale_fill_manual(values=pal, name="geoLocAdmin2") +
		scale_size_continuous(range=c(0.6,3), guide="none") +
		scale_x_continuous(breaks=mb+off, labels=ml, limits=c(min(ci[1]+off, rootx)-0.02, max(sts)+off+0.02)) +
		labs(title=title,
		     subtitle=glue("{nt} genomes | {phy$Nnode} internal nodes | MRCA {format(as.Date(date_decimal(root_date)),'%d %b %Y')} | {ntrans} parsimony geographic transitions")) +
		theme_tree2() +
		theme(plot.title=element_text(face="bold",size=12), plot.subtitle=element_text(size=8.5,color="grey40"),
		      axis.text.x=element_text(size=9,color="grey40"), legend.position=c(0.12,0.75),
		      legend.background=element_rect(fill="white",color="grey85"), legend.key.size=unit(0.9,"lines"))
}

td1_collapsed <- collapse_dated(td1, aln_len, min_subs = 0.5)
pg_coll <- geo_tree_fig(td1_collapsed, td1_collapsed$root.time, "")
print(pg_coll)
```

---